Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-07 paper/tidy-missing-data-paper.Rmd
cutting down on words + re-arranging
- verbs --> "missing data operations"
- move "missing data operations" to just after "data structures"
- adjust some heading levels

```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 4, fig.width = 4, out.width = "100%", warning = FALSE, fig.show = 'hold'}

@@

While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs challenging to integrate into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.

The analyst must
explore missing data using visualizations, summaries, and exploratory analyses,
but there is a lack of modern tools to efficiently explore missing data.

@@

With tidy data come tidy tools, which have the same input and output: tidy data. This consistency means multiple tools can be composed together in sequence, allowing for rapid, elegant operations. Contrasting tidy tools are messy tools. These have tidy input but messy output. Messy tools slow down analysis by shifting the focus from analysis to transforming output so it is the right shape for the next step in the analysis. This makes the work at each step harder to predict, and more complex and difficult to maintain. This disrupts workflow, and incites errors.

Tidy tools fall into three broad categories: data manipulation, visualization, and modelling. These are now discussed in turn.

@@

Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@tidyverse]. These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles and create fluid workflows for new domains. For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].

@@

### Coordinated names {#coordinated-names}

@@

### Missing value labels {#missing-labels}

@@

### Special missings {#special-missings}

@@

### Nabular data {#nabular-data}

@@

@@

### Imputing values below the range {#imp-below-range}

To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows the number of missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

@@

### Univariate split by missingness {#vis-split-by-miss}

Missingness of one variable can also be used to display different distributions in another. Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing. Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density. This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.

@@

To visualize missing values in two dimensions the missing values can be placed in the plot margins. This is achieved by imputing values below the range of the data. As in Section \@ref(imp-below-range), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring have been been combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)), where there is a mostly uniform spread of missing values for Solar.R and Ozone.

@@

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the \pkg{naniar} package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

@@

The overall number of missing values in a dataset is shown with `n_miss`, and the proportion or percent missings, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These be extended to summarize the number/amount of variables and cases containing a missing values by appending `_case` or `_var` to any of these summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.

@@

knitr::kable(
  table_of_fun, 
  digits = 2,
  caption = "Single number summaries of missing and complete data, applied to the airquality dataset. These functions follow consistent naming, to make them easy to remember, and to make their use clear.")
  # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

## Verbs {#verbs}

### scan: Searching for common missing value labels  {#verbs-search}

There are many common values for "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count). All occurrences of -99 can be searched for using the code:

knitr::kable(dat_ms, 
             caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

```{r miss-scan-print, eval = FALSE, echo = TRUE}
miss_scan_count(data = dat_ms, search = -99)
```

This returns a table of the occurrences of that value in each variable (Table \@ref(tab:miss-scan-count)). A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers. These contain values like "n/a", "na ", and ".", and -9, -99.

```{r miss-scan-count}
miss_scan_count(data = dat_ms, 
                search = -99) %>% 
  knitr::kable(
    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

### replace: Replacing missing value label with another {#verbs-replace-with}

Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values, using `replace_with_na`. This is useful when there is certainty of which values mean missing. For example "N/A", "N A", "Not Available", and "missing". Using the previous toy dataset, `dat_ms`, the value -99, or both -99 and 98 can be replaced with NA as follows:

```{r replace-with-na-code-chunk, echo = TRUE}
replace_with_na(dat_ms, replace = list(x = -99))

replace_with_na(dat_ms, replace = list(x = c(-99, -98)))
```

(**considering removing these examples**)

For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`. The syntax for these is slightly different to `replace_with_na`, but powerful:

```{r replace-with-na-scoped-variants, echo = TRUE}
replace_with_na_at(data = dat_ms,
                   .vars = "x",
                   condition = ~.x == -99)

replace_with_na_if(data = dat_ms,
                 .predicate = is.character,
                 condition = ~.x == "N/A")

replace_with_na_all(data = dat_ms,
                    condition = ~.x == -99)

### add: Adding missingness summary variables {#verbs-add-cols}

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively.

An example use of these features is in @Tierney2015,
where the proportion, number, or cluster of missings is used as the outcome in a
model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`:

```{r add-missing-info, echo = TRUE}
add_n_miss(dat_ms)
add_prop_miss(dat_ms)
add_miss_cluster(dat_ms)
```

### shadow: Creating nabular data {#verbs-nabular}

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `nabular()` or `bind_shadow()`. The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe. An example use of `nabular()` on the example data, `dat_ms` is shown below:

### flag: Describing different types of  missing values {#verbs-recode}

### impute: Imputing values {#verbs-impute}

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.

The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`. Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`). The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame. An example usage is shown below:

```{r impute-vector, echo = TRUE}

One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

### track: Shadow and impute missing values {#verbs-track}

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).

Imputed values can also be compared to complete case data by grouping by `any_missing`, and then summarizing (Table \@ref(tab:impute-summary)).

```r
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max))
```

```{r impute-summary}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max)) %>%
  knitr::kable(
    digits = 1, 
    caption = "Summary statistics of the previously imputed and complete cases. The mean and median values are similar, but the minimum and maximum values are different. The comparison of imputed values is similar to other dplyr summary workflows.")
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

@@

Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)). There are three main clusters of missing data: at the top there are many variables missing together; in the middle, building price and year built missing together, at the bottom, building area, year built, and land size are missing.

@@

Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss). 
Two patterns stand out: two, and five variables missing. This provides further evidence of the patterns of missingness seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

@@

\newpage

>>>>

```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}

@@

While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs challenging to integrate into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline. The analyst must
explore missing data using visualizations, summaries, and exploratory analyses,
but there is a lack of modern tools to efficiently explore missing data.

@@

With tidy data come tidy tools, which have the same input and output: tidy data. This consistency means multiple tools can be composed together in sequence, allowing for rapid, elegant operations. Contrasting tidy tools are messy tools. These have tidy input but messy output. Messy tools slow down analysis by shifting the focus from analysis to transforming output so it is the right shape for the next step in the analysis. This makes the work at each step harder to predict, and more complex and difficult to maintain. This disrupts workflow, and incites errors. Tidy tools fall into three broad categories: data manipulation, visualization, and modelling. These are now discussed in turn.

@@

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the "tidyverse" [@tidyverse]. These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles and create fluid workflows for new domains. For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].

@@

## Coordinated names {#coordinated-names}

@@

## Missing value labels {#missing-labels}

@@

## Special missings {#special-missings}

@@

## Nabular data {#nabular-data}

@@

# Missing data operations {#verbs}

## scan: Searching for common missing value labels  {#verbs-search}

There are many common values for "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, all occurrences of -99 can be searched for with `miss_scan_count(data, search = -99)`. This returns a table of the occurrences of that value in each variable.

# knitr::kable(dat_ms, 
#              caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

```{r miss-scan-count, eval = FALSE}
miss_scan_count(data = dat_ms, 
                search = -99) %>% 
  knitr::kable(
    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers. These contain values like "n/a", "na ", and ".", and -9, -99. These can be used with `miss_scan_count` to help identify common missing values.

## replace: Replacing missing value label with another {#verbs-replace-with}

Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values. This can be done with `replace_with_na()`. This is useful when there is certainty of which values mean missing. For example "N/A", "N A", "Not Available", and "missing". 
A dataset could have the values -99 in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))`

For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.

## add: Adding missingness summary variables {#verbs-add-cols}

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).

```{r add-missing-info}

tibble::tibble(
  `Function` = c(
    "add_n_miss(data)",
    "add_prop_miss(data)",
    "add_miss_cluster(data)"
    ),
  `Adds column which:` = c(
    "Contains the number missing values in a row",
    "Contains the proportion of missing values in a row",
    "Contains the missing value cluster"
  )) %>% 
    knitr::kable(caption = "Overview of the 'add' functions in naniar")

An example use of these features is in @Tierney2015,
where the proportion, number, or cluster of missings is used as the outcome in a
model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`.

## shadow: Creating nabular data {#verbs-nabular}

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `nabular()` or `bind_shadow()`. (The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe). An example use of `nabular()` some example data set is shown below:

## flag: Describing different types of  missing values {#verbs-recode}

## impute: Imputing values {#verbs-impute}

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis. The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`. Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`). The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame. One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

```{r impute-vector, eval = FALSE}

## track: Shadow and impute missing values {#verbs-track}

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows. This shows mean and median values are similar, but the minimum and maximum values are different.

```{r impute-summary, echo = TRUE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max)) 
```

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows the number of missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

@@

**Univariate split by missingness**. Missingness of one variable can also be used to display different distributions in another. Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing. Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density. This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.

@@

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data. As in Section \@ref(univariate-graphics), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring are combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)). There is a mostly uniform spread of missing values for Solar.R and Ozone.

@@

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the \pkg{naniar} package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

@@

The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These can summarize missings for variables and cases by appending `_case` or `_var` to any of the summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset 'airquality'.

@@

knitr::kable(
  table_of_fun, 
  digits = 2,
  caption = "Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.")
  # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

@@

Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)). There are three main clusters of missing data: at the top many variables go missing together; in the middle, building price and year built, at the bottom, building area, year built, and land size.

@@

Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss)). Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

@@
